\vspace{-0.5in}{\bfseries\scshape Transfer Learning for computationally limited training of Vision Transformers on art classification tasks } \\ \vspace{1ex} \normalsize{Bachelor's Project Thesis}\vspace{-2ex}

### 2022-08-07 writing/tex/experiments.tex
Incroporated some changes from Iris' feedback. NOW IT IS DONE. GOING TO HAND IT IN NEXTgit add -A!

\begin{table*}[tb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{lllllll}
\hline
\textbf{Model} & \textbf{Type} &  & \textbf{Material} & & \textbf{Artist} & \\
& \textbf{Accuracy} & \textbf{Bal. accuracy} & \textbf{Accuracy} & \textbf{Bal. accuracy} & \textbf{Accuracy} & \textbf{Bal. accuracy} \\ \hline
\textbf{vit\_b\_16} & 86.06\% $_{\pm 1.06\%}$ & 84.13\% $_{\pm 1.57\%}$ & 81.78\% $_{\pm 0.48\%}$ & 67.38\% $_{\pm 1.37\%}$ & 84.89\% $_{\pm 0.46\%}$ & 81.42\% $_{\pm 0.42\%}$  \\
\textbf{swin\_b} & \cellcolor{bestcol}89.43\% $_{\pm 0.93\%}$ & \cellcolor{bestcol}87.47\% $_{\pm 1.02\%}$ & \cellcolor{bestcol}85.87\% $_{\pm 0.35\%}$ & \cellcolor{bestcol}71.19\% $_{\pm 1.60\%}$ & \cellcolor{bestcol}90.40\% $_{\pm 0.65\%}$ & \cellcolor{bestcol}88.64\% $_{\pm 0.78\%}$  \\
\textbf{beit\_b\_16} & \cellcolor{worstcol}82.26\% $_{\pm 0.72\%}$ & \cellcolor{worstcol}77.75\% $_{\pm 0.27\%}$ & 76.87\% $_{\pm 0.96\%}$ & 60.16\% $_{\pm 1.56\%}$ & 79.70\% $_{\pm 0.69\%}$ & 75.35\% $_{\pm 1.09\%}$  \\
\textbf{deit\_b\_16} & 88.18\% $_{\pm 0.66\%}$ & 85.36\% $_{\pm 0.54\%}$ & 82.80\% $_{\pm 1.12\%}$ & 66.46\% $_{\pm 1.03\%}$ & 88.13\% $_{\pm 0.76\%}$ & 85.62\% $_{\pm 0.87\%}$  \\\hdashline
\textbf{vgg19} & 83.93\% $_{\pm 0.72\%}$ & 83.35\% $_{\pm 0.81\%}$ & 76.87\% $_{\pm 0.44\%}$ & 61.39\% $_{\pm 1.47\%}$ & 82.01\% $_{\pm 0.66\%}$ & 78.10\% $_{\pm 0.77\%}$  \\
\textbf{resnet50} & 85.51\% $_{\pm 0.64\%}$ & 82.33\% $_{\pm 1.85\%}$ & 80.99\% $_{\pm 0.82\%}$ & 65.51\% $_{\pm 0.93\%}$ & 87.71\% $_{\pm 1.06\%}$ & 85.12\% $_{\pm 1.34\%}$  \\
\textbf{eff. netv2\_m} & 83.41\% $_{\pm 0.76\%}$ & 82.05\% $_{\pm 1.25\%}$  & \cellcolor{worstcol}75.96\% $_{\pm 1.24\%}$ & \cellcolor{worstcol}59.15\% $_{\pm 1.24\%}$ & \cellcolor{worstcol}78.62\% $_{\pm 1.07\%}$ & \cellcolor{worstcol}73.92\% $_{\pm 0.96\%}$  \\
\textbf{convnext\_b} & \cellcolor{secondbestcol}89.19\% $_{\pm 0.64\%}$ & \cellcolor{secondbestcol}86.95\% $_{\pm 1.38\%}$ & \cellcolor{secondbestcol}84.14\% $_{\pm 0.92\%}$ & \cellcolor{secondbestcol}69.10\% $_{\pm 1.05\%}$ & \cellcolor{secondbestcol}90.13\% $_{\pm 0.94\%}$ & \cellcolor{secondbestcol}87.84\% $_{\pm 1.07\%}$  \\\hline
\end{tabular}
}
\caption{Testing performance after off-the-shelf learning. Results are similar to validation accuracy shown in figure \ref{results:img:ots}, with \texttt{Swin} and \texttt{ConvNext} showing the respective best and second best performance in all cases.}
\label{results:tab:ots}
\end{table*}

Table \ref{results:tab:ots} shows the testing performance of OTS-trained models on the 3 classification tasks, while figure \ref{results:img:ots} shows corresponding validation accuracies. Performance on the testing set is comparable among all tasks in terms of rankings. Other observations are consistent as well, such as \texttt{ConvNext} taking the most epochs to converge.

@@

\begin{figure*}[tb]
    \centering
    \def\svgwidth{\textwidth}
    \input{img/ft_mean_accuracy.pdf_tex}
    \caption{Average validation accuracy when fine-tuning models on the target task. Overall, performance is higher compared to off-the-shelf learning in figure \ref{results:img:ots}, while the differences between models are also smaller here. The \texttt{Swin} VT and \texttt{ConvNext} CNN remain the best performing models, while in general, results are less favorable for VTs than they were after off-the-shelf learning.}
    \label{results:img:ft}
\end{figure*}

\begin{table*}[tb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{lllllll}
\hline
\textbf{Model} & \textbf{Type} &  & \textbf{Material} & & \textbf{Artist} & \\
& \textbf{Accuracy} & \textbf{Bal. accuracy} & \textbf{Accuracy} & \textbf{Bal. accuracy} & \textbf{Accuracy} & \textbf{Bal. accuracy} \\ \hline
\textbf{vit\_b\_16} & \cellcolor{worstcol}90.11\% $_{\pm 0.35\%}$ & 87.40\% $_{\pm 0.29\%}$ & 87.42\% $_{\pm 0.45\%}$ & 73.46\% $_{\pm 1.44\%}$ & 92.05\% $_{\pm 0.44\%}$ & 89.77\% $_{\pm 0.38\%}$  \\
\textbf{swin\_b} & \cellcolor{bestcol}92.17\% $_{\pm 0.98\%}$ & \cellcolor{secondbestcol}89.71\% $_{\pm 1.03\%}$ & \cellcolor{bestcol}89.35\% $_{\pm 0.68\%}$ & 77.16\% $_{\pm 2.98\%}$ & \cellcolor{bestcol}95.05\% $_{\pm 0.47\%}$ & \cellcolor{bestcol}93.94\% $_{\pm 0.81\%}$  \\
\textbf{beit\_b\_16} & 90.81\% $_{\pm 0.41\%}$ & 87.95\% $_{\pm 0.69\%}$ & \cellcolor{worstcol}85.74\% $_{\pm 0.37\%}$ & \cellcolor{worstcol}72.12\% $_{\pm 1.38\%}$ & \cellcolor{worstcol}91.27\% $_{\pm 1.13\%}$ & \cellcolor{worstcol}88.83\% $_{\pm 1.63\%}$  \\
\textbf{deit\_b\_16} & 91.78\% $_{\pm 0.64\%}$ & 89.22\% $_{\pm 0.90\%}$ & 87.85\% $_{\pm 1.12\%}$ & 74.42\% $_{\pm 1.99\%}$ & 93.37\% $_{\pm 1.15\%}$ & 91.67\% $_{\pm 1.55\%}$  \\\hdashline
\textbf{vgg19} & 90.54\% $_{\pm 0.37\%}$ & \cellcolor{worstcol}87.05\% $_{\pm 1.03\%}$ & 85.74\% $_{\pm 1.40\%}$ & 72.43\% $_{\pm 3.03\%}$ & 92.20\% $_{\pm 0.49\%}$ & 90.18\% $_{\pm 0.72\%}$  \\
\textbf{resnet50} & 91.78\% $_{\pm 0.44\%}$ & 88.24\% $_{\pm 0.59\%}$ & 88.69\% $_{\pm 0.99\%}$ & \cellcolor{secondbestcol}77.97\% $_{\pm 2.25\%}$ & \cellcolor{secondbestcol}94.72\% $_{\pm 0.74\%}$ & \cellcolor{secondbestcol}93.41\% $_{\pm 1.05\%}$  \\
\textbf{eff. netv2\_m} & 90.87\% $_{\pm 0.67\%}$ & 88.34\% $_{\pm 1.37\%}$ & 87.55\% $_{\pm 1.15\%}$ & 75.31\% $_{\pm 1.60\%}$ & 92.65\% $_{\pm 0.54\%}$ & 90.84\% $_{\pm 0.51\%}$  \\
\textbf{convnext\_b} & \cellcolor{secondbestcol}92.15\% $_{\pm 0.40\%}$ & \cellcolor{bestcol}89.82\% $_{\pm 1.18\%}$ & \cellcolor{secondbestcol}88.79\% $_{\pm 1.07\%}$ & \cellcolor{bestcol}78.40\% $_{\pm 1.26\%}$ & 94.60\% $_{\pm 0.54\%}$ & 93.13\% $_{\pm 0.61\%}$  \\\hline
\end{tabular}
}
\caption{Testing performance after fine-tuning. Results are again similar to validation accuracy in figure \ref{results:img:ft}. Compared to table \ref{results:tab:ots}, it is striking that the \texttt{ConvNext} CNN now often takes the lead with respect to balanced accuracy, and that the \texttt{ResNet50} CNN often takes second place.}
\label{results:tab:ft}
\end{table*}

Once again, table \ref{results:tab:ft} shows testing performance, while figure \ref{results:img:ft} shows validation accuracies.

@@

\paragraph{How do these results compare to related studies?}
Of all related work mentioned in section \ref{related_work}, \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}) was the only one that examined VTs' OTS learning properties. It reported the best OTS learning performance for the \texttt{Swin} VT, which is much in line with the results in table \ref{results:tab:ots}.

Besides \texttt{Swin}, \citeauthor{zhou2021convnets} also included \texttt{ViT} and \texttt{ResNet101} in its comparisons (among others), and consistently found that \texttt{ResNet101} outperformed \texttt{ViT}. The current study, on the other hand, uses \texttt{ResNet50}, which performs worse than \texttt{ViT} on all tasks except \textit{Artist classification}. Running the same experiments with \texttt{ResNet101} does, however, lead to similar findings as \citeauthor{zhou2021convnets}. On \textit{Type classification}, for example, it achieves a mean testing accuracy of  $86.21\% (\pm 0.70\%)$, which is indeed slightly higher than \texttt{ViT}'s 86.06\% shown in table \ref{results:tab:ots}. Mind that these differences are slight compared to the 89.43\% accuracy achieved by \texttt{Swin} here, so replacing \texttt{ResNet50} with \texttt{ResNet101} would not have affected the overall observations that much.

Finally, because \texttt{VGG19} was specifically selected for its good OTS performance in \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}), it is interesting that this model is outperformed by \texttt{ResNet50} (which that paper also included) in table \ref{results:tab:ots}. \texttt{VGG19} is the largest model of all, so this discrepancy could be an indication that the smaller datasets used in the current study are too small for \texttt{VGG19}. Additionally, it might also be attributable to the smaller number of classes used here: \texttt{VGG19} has the largest final layer, containing 4096 inputs. While in \citeauthor{sabatelli2018deep} this might have provided the expressive power needed to differentiate between a few hundred classes after just OTS learning, here it likely causes overfitting before good convergence is reached.

@@

A hint of an answer can, however, be found when plotting activation of successively deeper attention layers, as is done in figure \ref{results:img:layers}. While earlier layers retain much spatial information, later ones show more sporadic activation patterns. This is what one might expect from a hierarchical feature extractor: the random pixels lighting up might, for example, correspond to specific, high-level properties of the input. Nevertheless, it should be emphasized that no strong conclusions should be drawn from these images, and that more research with respect to explainability is desired.
% Also mention that this was unexpected (but in line with blabla);
% Feature extractor/hierarchical representation building
% Layers img.
% Maybe mention that image specific inductive bias isn't a problem for swin? TODO somewhere else!

@@

\paragraph{How do these results compare to related studies?}
That the \texttt{Swin} VT performs well after FT, is again consistent with \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}). In that paper, however, \texttt{ViT} outperforms the \texttt{ResNet101} and \texttt{ResNet152} CNNs. This differs from the results in table \ref{results:tab:ft}, where \texttt{ViT} does not perform well overall, and even shows the worst testing accuracy on \textit{Type classification}, with 90.11\%. \texttt{ResNet50} achieves 91.78\% accuracy here, and testing the same \texttt{ResNet101} and \texttt{ResNet152} versions that \citeauthor{zhou2021convnets} used does not change much, as this leads to 91.96\% ($\pm$ 0.59\%) and 91.94\% ($\pm$ 0.25) accuracy, respectively.

Results are also less favorable for VTs compared to \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}). In that paper, the `tiny' version of \texttt{DeiT} outperforms \texttt{ResNet50}, while in table \ref{results:tab:ft} this same \texttt{ResNet50} performs slightly better than the `base' version of \texttt{DeiT}. Going `tiny' also does not help: on \textit{Type classification}, for instance, `tiny' \texttt{DeiT} gets 90.44\% ($\pm$ 0.93\%), which is even worse than `base' \texttt{DeiT}'s 91.78\% in table \ref{results:tab:ots}.

@@

\begin{figure*}[tb]
    \centering
    \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=8cm]{img/img011_salience.png}
    \caption{Picture}
    \label{results:img:sal:picture}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=8cm]{img/img005_salience.png}
    \caption{Dish}
    \label{results:img:sal:dish}
    \end{subfigure}
    \caption{Saliency maps for \texttt{ConvNext}, \texttt{DeiT}, and \texttt{Swin} after both off-the-shelf learning (OTS), and fine-tuning (FT). \texttt{ConvNext} appears to benefit more when training the whole model instead of the last layer, because it shows the biggest differences between OTS learning and FT. In the case of image \ref{results:img:sal:dish}, this even makes the difference between correct and incorrect classification.}
    \label{results:img:sal}
\end{figure*}

VTs and CNNs are much closer together after FT in table \ref{results:tab:ft} than after OTS learning in table \ref{results:tab:ots}. It is as if OTS learning already unlocks much of VTs' potential, while CNNs need an FT learning scheme for this.

Supporting `evidence' might be visible in figure \ref{results:img:sal}. It plots saliency maps for one CNN (\texttt{ConvNext}), and two VTs (\texttt{DeiT} and \texttt{Swin}). It is interesting that \texttt{ConvNext} shows much bigger differences between the OTS and FT plots. In figure \ref{results:img:sal:picture}, for example, it learns to just look at the picture frame after FT. In figure \ref{results:img:sal:dish} it learns to look more at the center, which allows it to correctly classify the input as a dish instead of a plate\footnote{No idea what the difference is between these classes, but these models do seem to know, so that is good!}. % TODO Maybe delete this comment, not very professional
These bigger differences for \texttt{ConvNext} align well with observations made earlier, as they suggest that there is still much improvement possible after just OTS learning for CNNs.

The images in figure \ref{results:img:sal} are representative examples, and show observations that could have been made from other images too. How representative the 3 models are for all 8 is debatable, and was not tested. In general, a big disclaimer should be added to these results, because some hand waving is involved. Importantly, 3 different salience mapping techniques were used: GradCam \citep{selvaraju2017grad} was implemented for \texttt{ConvNext}, while Attention Rollout \citep{abnar2020quantifying} was implemented for \texttt{DeiT}. This already makes the comparison somewhat unfair, and it becomes even worse when \texttt{Swin} is added, for which a self-invented method was used. This method is similar to GradCam, but instead of multiplying \textit{average} CNN channel gradients with said channel's activation, it directly multiplies activation with corresponding gradients in the output of one of the later transformer blocks\footnote{For the full implementation of all techniques, one is referred to the source code on: \\ \url{https://github.com/IndoorAdventurer/ViTTransferLearningForArtClassification}.}. Despite this, saliency maps were included, as they are interesting nonetheless.
% Show saliency maps here
% MENTION TECHNIQUES AND SHORTCOMMINGS!!!

@@

Then there is the problem of balance. It was mentioned earlier that balanced accuracy falls behind plain accuracy the most for \textit{Material classification}, and the least for \textit{Artist classification}. At the same time, table \ref{methods:datasets} shows that it is also \textit{Material classification} which is the least balanced (Gini-coefficient of 0.563), while it is \textit{Artist classification} that is the most balanced (coefficient 0.236). This suggests that balance has a positive effect on testing accuracy, from which \textit{Artist classification}, in that case, benefits the most.

@@

Of course, final accuracy is not the full story, since time/accuracy trade-offs, for example, can also determine which model is preferred. Figure \ref{results:img:ots_vs_ft_type} plots the \textit{Type classification} accuracies of tables \ref{results:tab:ots} and \ref{results:tab:ft} against training time per epoch. Total training time was not chosen instead, as this measure depends much more on selected hyperparameters.

@@

All `tiny' VTs are around the same size as \texttt{ResNet50}, and using these would have changed the situation completely. As such, taking figure \ref{results:img:ots_vs_ft_type} with a grain of salt would be appropriate.

>>>>

\begin{table*}[tb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{lllllll}
\hline
\textbf{Model} & \textbf{Type} &  & \textbf{Material} & & \textbf{Artist} & \\
& \textbf{Accuracy} & \textbf{Bal. accuracy} & \textbf{Accuracy} & \textbf{Bal. accuracy} & \textbf{Accuracy} & \textbf{Bal. accuracy} \\ \hline
\textbf{vit\_b\_16} & 86.06\% $_{\pm 1.06\%}$ & 84.13\% $_{\pm 1.57\%}$ & 81.78\% $_{\pm 0.48\%}$ & 67.38\% $_{\pm 1.37\%}$ & 84.89\% $_{\pm 0.46\%}$ & 81.42\% $_{\pm 0.42\%}$  \\
\textbf{swin\_b} & \cellcolor{bestcol}89.43\% $_{\pm 0.93\%}$ & \cellcolor{bestcol}87.47\% $_{\pm 1.02\%}$ & \cellcolor{bestcol}85.87\% $_{\pm 0.35\%}$ & \cellcolor{bestcol}71.19\% $_{\pm 1.60\%}$ & \cellcolor{bestcol}90.40\% $_{\pm 0.65\%}$ & \cellcolor{bestcol}88.64\% $_{\pm 0.78\%}$  \\
\textbf{beit\_b\_16} & \cellcolor{worstcol}82.26\% $_{\pm 0.72\%}$ & \cellcolor{worstcol}77.75\% $_{\pm 0.27\%}$ & 76.87\% $_{\pm 0.96\%}$ & 60.16\% $_{\pm 1.56\%}$ & 79.70\% $_{\pm 0.69\%}$ & 75.35\% $_{\pm 1.09\%}$  \\
\textbf{deit\_b\_16} & 88.18\% $_{\pm 0.66\%}$ & 85.36\% $_{\pm 0.54\%}$ & 82.80\% $_{\pm 1.12\%}$ & 66.46\% $_{\pm 1.03\%}$ & 88.13\% $_{\pm 0.76\%}$ & 85.62\% $_{\pm 0.87\%}$  \\\hdashline
\textbf{vgg19} & 83.93\% $_{\pm 0.72\%}$ & 83.35\% $_{\pm 0.81\%}$ & 76.87\% $_{\pm 0.44\%}$ & 61.39\% $_{\pm 1.47\%}$ & 82.01\% $_{\pm 0.66\%}$ & 78.10\% $_{\pm 0.77\%}$  \\
\textbf{resnet50} & 85.51\% $_{\pm 0.64\%}$ & 82.33\% $_{\pm 1.85\%}$ & 80.99\% $_{\pm 0.82\%}$ & 65.51\% $_{\pm 0.93\%}$ & 87.71\% $_{\pm 1.06\%}$ & 85.12\% $_{\pm 1.34\%}$  \\
\textbf{eff. netv2\_m} & 83.41\% $_{\pm 0.76\%}$ & 82.05\% $_{\pm 1.25\%}$  & \cellcolor{worstcol}75.96\% $_{\pm 1.24\%}$ & \cellcolor{worstcol}59.15\% $_{\pm 1.24\%}$ & \cellcolor{worstcol}78.62\% $_{\pm 1.07\%}$ & \cellcolor{worstcol}73.92\% $_{\pm 0.96\%}$  \\
\textbf{convnext\_b} & \cellcolor{secondbestcol}89.19\% $_{\pm 0.64\%}$ & \cellcolor{secondbestcol}86.95\% $_{\pm 1.38\%}$ & \cellcolor{secondbestcol}84.14\% $_{\pm 0.92\%}$ & \cellcolor{secondbestcol}69.10\% $_{\pm 1.05\%}$ & \cellcolor{secondbestcol}90.13\% $_{\pm 0.94\%}$ & \cellcolor{secondbestcol}87.84\% $_{\pm 1.07\%}$  \\\hline
\end{tabular}
}
\caption{Testing performance after off-the-shelf learning. Results are similar to validation accuracy shown in Figure \ref{results:img:ots}, with \texttt{Swin} and \texttt{ConvNext} showing the respective best and second best performance in all cases.}
\label{results:tab:ots}
\end{table*}

Table \ref{results:tab:ots} shows the testing performance of OTS-trained models on the 3 classification tasks, while Figure \ref{results:img:ots} shows corresponding validation accuracies. Performance on the testing set is comparable among all tasks in terms of rankings. Other observations are consistent as well, such as \texttt{ConvNext} taking the most epochs to converge.

@@

\begin{figure*}[tb]
    \centering
    \def\svgwidth{\textwidth}
    \input{img/ft_mean_accuracy.pdf_tex}
    \caption{Average validation accuracy when fine-tuning models on the target task. Overall, performance is higher compared to off-the-shelf learning in Figure \ref{results:img:ots}, while the differences between models are also smaller here. The \texttt{Swin} VT and \texttt{ConvNext} CNN remain the best performing models, while in general, results are less favorable for VTs than they were after off-the-shelf learning.}
    \label{results:img:ft}
\end{figure*}

\begin{table*}[tb]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{lllllll}
\hline
\textbf{Model} & \textbf{Type} &  & \textbf{Material} & & \textbf{Artist} & \\
& \textbf{Accuracy} & \textbf{Bal. accuracy} & \textbf{Accuracy} & \textbf{Bal. accuracy} & \textbf{Accuracy} & \textbf{Bal. accuracy} \\ \hline
\textbf{vit\_b\_16} & \cellcolor{worstcol}90.11\% $_{\pm 0.35\%}$ & 87.40\% $_{\pm 0.29\%}$ & 87.42\% $_{\pm 0.45\%}$ & 73.46\% $_{\pm 1.44\%}$ & 92.05\% $_{\pm 0.44\%}$ & 89.77\% $_{\pm 0.38\%}$  \\
\textbf{swin\_b} & \cellcolor{bestcol}92.17\% $_{\pm 0.98\%}$ & \cellcolor{secondbestcol}89.71\% $_{\pm 1.03\%}$ & \cellcolor{bestcol}89.35\% $_{\pm 0.68\%}$ & 77.16\% $_{\pm 2.98\%}$ & \cellcolor{bestcol}95.05\% $_{\pm 0.47\%}$ & \cellcolor{bestcol}93.94\% $_{\pm 0.81\%}$  \\
\textbf{beit\_b\_16} & 90.81\% $_{\pm 0.41\%}$ & 87.95\% $_{\pm 0.69\%}$ & \cellcolor{worstcol}85.74\% $_{\pm 0.37\%}$ & \cellcolor{worstcol}72.12\% $_{\pm 1.38\%}$ & \cellcolor{worstcol}91.27\% $_{\pm 1.13\%}$ & \cellcolor{worstcol}88.83\% $_{\pm 1.63\%}$  \\
\textbf{deit\_b\_16} & 91.78\% $_{\pm 0.64\%}$ & 89.22\% $_{\pm 0.90\%}$ & 87.85\% $_{\pm 1.12\%}$ & 74.42\% $_{\pm 1.99\%}$ & 93.37\% $_{\pm 1.15\%}$ & 91.67\% $_{\pm 1.55\%}$  \\\hdashline
\textbf{vgg19} & 90.54\% $_{\pm 0.37\%}$ & \cellcolor{worstcol}87.05\% $_{\pm 1.03\%}$ & 85.74\% $_{\pm 1.40\%}$ & 72.43\% $_{\pm 3.03\%}$ & 92.20\% $_{\pm 0.49\%}$ & 90.18\% $_{\pm 0.72\%}$  \\
\textbf{resnet50} & 91.78\% $_{\pm 0.44\%}$ & 88.24\% $_{\pm 0.59\%}$ & 88.69\% $_{\pm 0.99\%}$ & \cellcolor{secondbestcol}77.97\% $_{\pm 2.25\%}$ & \cellcolor{secondbestcol}94.72\% $_{\pm 0.74\%}$ & \cellcolor{secondbestcol}93.41\% $_{\pm 1.05\%}$  \\
\textbf{eff. netv2\_m} & 90.87\% $_{\pm 0.67\%}$ & 88.34\% $_{\pm 1.37\%}$ & 87.55\% $_{\pm 1.15\%}$ & 75.31\% $_{\pm 1.60\%}$ & 92.65\% $_{\pm 0.54\%}$ & 90.84\% $_{\pm 0.51\%}$  \\
\textbf{convnext\_b} & \cellcolor{secondbestcol}92.15\% $_{\pm 0.40\%}$ & \cellcolor{bestcol}89.82\% $_{\pm 1.18\%}$ & \cellcolor{secondbestcol}88.79\% $_{\pm 1.07\%}$ & \cellcolor{bestcol}78.40\% $_{\pm 1.26\%}$ & 94.60\% $_{\pm 0.54\%}$ & 93.13\% $_{\pm 0.61\%}$  \\\hline
\end{tabular}
}
\caption{Testing performance after fine-tuning. Results are again similar to validation accuracy in Figure \ref{results:img:ft}. Compared to Table \ref{results:tab:ots}, it is striking that the \texttt{ConvNext} CNN now often takes the lead with respect to balanced accuracy, and that the \texttt{ResNet50} CNN often takes second place.}
\label{results:tab:ft}
\end{table*}

Once again, Table \ref{results:tab:ft} shows testing performance, while Figure \ref{results:img:ft} shows validation accuracies.

@@

\paragraph{How do these results compare to related studies?}
Of all related work mentioned in section \ref{related_work}, \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}) was the only one that examined VTs' OTS learning properties. It reported the best OTS learning performance for the \texttt{Swin} VT, which is much in line with the results in Table \ref{results:tab:ots}.

Besides \texttt{Swin}, \citeauthor{zhou2021convnets} also included \texttt{ViT} and \texttt{ResNet101} in its comparisons (among others), and consistently found that \texttt{ResNet101} outperformed \texttt{ViT}. The current study, on the other hand, uses \texttt{ResNet50}, which performs worse than \texttt{ViT} on all tasks except \textit{Artist classification}. Running the same experiments with \texttt{ResNet101} does, however, lead to similar findings as \citeauthor{zhou2021convnets}. On \textit{Type classification}, for example, it achieves a mean testing accuracy of  $86.21\% (\pm 0.70\%)$, which is indeed slightly higher than \texttt{ViT}'s 86.06\% shown in Table \ref{results:tab:ots}. Mind that these differences are slight compared to the 89.43\% accuracy achieved by \texttt{Swin} here, so replacing \texttt{ResNet50} with \texttt{ResNet101} would not have affected the overall observations that much.

Finally, because \texttt{VGG19} was specifically selected for its good OTS performance in \citeauthor{sabatelli2018deep} (\citeyear{sabatelli2018deep}), it is interesting that this model is outperformed by \texttt{ResNet50} (which that paper also included) in Table \ref{results:tab:ots}. \texttt{VGG19} is the largest model of all, so this discrepancy could be an indication that the smaller datasets used in the current study are too small for \texttt{VGG19}. Additionally, it might also be attributable to the smaller number of classes used here: \texttt{VGG19} has the largest final layer, containing 4096 inputs. While in \citeauthor{sabatelli2018deep} this might have provided the expressive power needed to differentiate between a few hundred classes after just OTS learning, here it likely causes overfitting before good convergence is reached.

@@

A hint of an answer can, however, be found when plotting activation of successively deeper attention layers, as is done in Figure \ref{results:img:layers}. While earlier layers retain much spatial information, later ones show more sporadic activation patterns. This is what one might expect from a hierarchical feature extractor: the random pixels lighting up might, for example, correspond to specific, high-level properties of the input. Nevertheless, it should be emphasized that no strong conclusions should be drawn from these images, and that more research with respect to explainability is desired.
% Also mention that this was unexpected (but in line with blabla);
% Feature extractor/hierarchical representation building
% Layers img.
% Maybe mention that image specific inductive bias isn't a problem for swin? TODO somewhere else!

@@

\paragraph{How do these results compare to related studies?}
That the \texttt{Swin} VT performs well after FT, is again consistent with \citeauthor{zhou2021convnets} (\citeyear{zhou2021convnets}). In that paper, however, \texttt{ViT} outperforms the \texttt{ResNet101} and \texttt{ResNet152} CNNs. This differs from the results in Table \ref{results:tab:ft}, where \texttt{ViT} does not perform well overall, and even shows the worst testing accuracy on \textit{Type classification}, with 90.11\%. \texttt{ResNet50} achieves 91.78\% accuracy here, and testing the same \texttt{ResNet101} and \texttt{ResNet152} versions that \citeauthor{zhou2021convnets} used does not change much, as this leads to 91.96\% ($\pm$ 0.59\%) and 91.94\% ($\pm$ 0.25) accuracy, respectively.

Results are also less favorable for VTs compared to \citeauthor{matsoukas2021time} (\citeyear{matsoukas2021time}). In that paper, the `tiny' version of \texttt{DeiT} outperforms \texttt{ResNet50}, while in Table \ref{results:tab:ft} this same \texttt{ResNet50} performs slightly better than the `base' version of \texttt{DeiT}. Going `tiny' also does not help: on \textit{Type classification}, for instance, `tiny' \texttt{DeiT} gets 90.44\% ($\pm$ 0.93\%), which is even worse than `base' \texttt{DeiT}'s 91.78\% in Table \ref{results:tab:ots}.

@@

\begin{figure*}[tb]
    \centering
    \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=8cm]{img/img011_salience.png}
    \caption{Picture}
    \label{results:img:sal:picture}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
    \includegraphics[width=8cm]{img/img005_salience.png}
    \caption{Dish}
    \label{results:img:sal:dish}
    \end{subfigure}
    \caption{Saliency maps for \texttt{ConvNext}, \texttt{DeiT}, and \texttt{Swin} after both off-the-shelf learning (OTS), and fine-tuning (FT). \texttt{ConvNext} appears to benefit more when training the whole model instead of the last layer, because it shows the biggest differences between OTS learning and FT. In the case of Figure \ref{results:img:sal:dish}, this even makes the difference between correct and incorrect classification.}
    \label{results:img:sal}
\end{figure*}

VTs and CNNs are much closer together after FT in Table \ref{results:tab:ft} than after OTS learning in Table \ref{results:tab:ots}. It is as if OTS learning already unlocks much of VTs' potential, while CNNs need an FT learning scheme for this.

Supporting `evidence' might be visible in Figure \ref{results:img:sal}. It plots saliency maps for one CNN (\texttt{ConvNext}), and two VTs (\texttt{DeiT} and \texttt{Swin}). It is interesting that \texttt{ConvNext} shows much bigger differences between the OTS and FT plots. In Figure \ref{results:img:sal:picture}, for example, it learns to just look at the picture frame after FT. In Figure \ref{results:img:sal:dish} it learns to look more at the center, which allows it to correctly classify the input as a dish instead of a plate\footnote{No idea what the difference is between these classes, but these models do seem to know, so that is good!}. % TODO Maybe delete this comment, not very professional
These bigger differences for \texttt{ConvNext} align well with observations made earlier, as they suggest that there is still much improvement possible after just OTS learning for CNNs.

The images in Figure \ref{results:img:sal} are representative examples, and show observations that could have been made from other images too. How representative the 3 models are for all 8 is debatable, and was not tested. In general, a big disclaimer should be added to these results, because some hand waving is involved. Importantly, 3 different salience mapping techniques were used: GradCam \citep{selvaraju2017grad} was implemented for \texttt{ConvNext}, while Attention Rollout \citep{abnar2020quantifying} was implemented for \texttt{DeiT}. This already makes the comparison somewhat unfair, and it becomes even worse when \texttt{Swin} is added, for which a self-invented method was used. This method is similar to GradCam, but instead of multiplying \textit{average} CNN channel gradients with said channel's activation, it directly multiplies activation with corresponding gradients in the output of one of the later transformer blocks\footnote{For the full implementation of all techniques, one is referred to the source code on: \\ \url{https://github.com/IndoorAdventurer/ViTTransferLearningForArtClassification}.}. Despite this, saliency maps were included, as they are interesting nonetheless.
% Show saliency maps here
% MENTION TECHNIQUES AND SHORTCOMMINGS!!!

@@

Then there is the problem of balance. It was mentioned earlier that balanced accuracy falls behind plain accuracy the most for \textit{Material classification}, and the least for \textit{Artist classification}. At the same time, Table \ref{methods:datasets} shows that it is also \textit{Material classification} which is the least balanced (Gini-coefficient of 0.563), while it is \textit{Artist classification} that is the most balanced (coefficient 0.236). This suggests that balance has a positive effect on testing accuracy, from which \textit{Artist classification}, in that case, benefits the most.

@@

Of course, final accuracy is not the full story, since time/accuracy trade-offs, for example, can also determine which model is preferred. Figure \ref{results:img:ots_vs_ft_type} plots the \textit{Type classification} accuracies of Tables \ref{results:tab:ots} and \ref{results:tab:ft} against training time per epoch. Total training time was not chosen instead, as this measure depends much more on selected hyperparameters.

@@

All `tiny' VTs are around the same size as \texttt{ResNet50}, and using these would have changed the situation completely. As such, taking Figure \ref{results:img:ots_vs_ft_type} with a grain of salt would be appropriate.
